---
title: "04_10_2024_statistical_analyses_septoria_in_a_bag"
output: html_document
date: "2024-04-10"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Packages

```{r}
library(vegan)
library(ggplot2)
library(dplyr)
library(ape)
library(tidyr)
library(ggpubr)
library(hillR)
library(rstatix)
library(car)
library(stringr)
library(FSA)
library(plotly)
```


# Generating Functions

Creating custom functions to be used for statistical analyses. 

## Rarefaction Curve Plotter

The below function takes the following objects to generate rarefactions curves generated in ggplot:

1. **nonrarefied_asv_tab**: The original non-rarefied asv table 
2. **metadata**: The original metadata object that describes origins of samples
3. **vars**: Factor used to determine the cut-off for the maximum sequencing depth that can be used to retain at least 3 replicates per factor condition. 
4. **color**: metadata factor used to color lines

The function will return a list contatining the following objects:

1. rarefaction_obj: A dataframe containing the rarefaction sampling information for each sample
2. rarefaction_plot: A ggplot depicting the rarefaction curves. 

```{r}
# Constructing rarefaction curves for fungal root communities
rarefaction_plotter<-function(nonrarefied_asv_tab,metadata,vars,color){
  # Constructing rarefaction curves for fungal root communities
  rarefaction_tidy<-rarecurve(nonrarefied_asv_tab,tidy = TRUE)
  rarefaction_obj<-merge(metadata,rarefaction_tidy,by.x=0,by.y="Site")
  colnames(rarefaction_obj)<-sub("Row.names","Sample.ID",colnames(rarefaction_obj))
  replicate_min<-c()
  for(i in 1:length(nonrarefied_asv_tab)){
    sample_depth<-sort(rowSums(nonrarefied_asv_tab))[i]
    rarefied_table<-suppressWarnings(as.data.frame(rrarefy(nonrarefied_asv_tab,sample=sample_depth)))
    ds.rarefied<-as.data.frame(subset(rarefied_table,rowSums(rarefied_table)>=sample_depth))
    test.df<-as.data.frame(merge(ds.rarefied,metadata,by.x=0,by.y=0)) %>%
  dplyr::count({{vars}},.drop=FALSE)
if((nrow(test.df[test.df$n<3,]))>0){
  replicate_min<-c(replicate_min,sample_depth)
  break
}
  }
  seq_counts_unrarefied<-sort(rowSums(nonrarefied_asv_tab))
  rarefaction_plot<-ggplot()+
  geom_line(data=rarefaction_obj,aes(x=Sample,y=Species,color=.data[[color]],group=.data[["Sample.ID"]]))+
  geom_vline(data=NULL,xintercept=c(seq_counts_unrarefied[c(round(length(seq_counts_unrarefied)*0.1),round(length(seq_counts_unrarefied)*0.30),round(length(seq_counts_unrarefied)*0.60),round(length(seq_counts_unrarefied)*0.9),length(seq_counts_unrarefied))]),linetype=2)+
  geom_vline(data=NULL,xintercept=replicate_min,color="red")+
  #geom_text(data=NULL,aes(x=c(rowSums(nonrarefied_asv_tab[rowSums(nonrarefied_asv_tab)<replicate_min,]))-1000,y=max(rarefaction_obj$Species+10),label=c(rowSums(nonrarefied_asv_tab[rowSums(nonrarefied_asv_tab)<replicate_min,])),angle=90))+
    scale_x_continuous(
      breaks=c(seq_counts_unrarefied[c(round(length(seq_counts_unrarefied)*0.1),round(length(seq_counts_unrarefied)*0.30),round(length(seq_counts_unrarefied)*0.60),round(length(seq_counts_unrarefied)*0.9),length(seq_counts_unrarefied))]),
      labels=c(seq_counts_unrarefied[c(round(length(seq_counts_unrarefied)*0.1),round(length(seq_counts_unrarefied)*0.3),round(length(seq_counts_unrarefied)*0.60),round(length(seq_counts_unrarefied)*0.9),length(seq_counts_unrarefied))]))+
  geom_text(data=NULL,aes(x=replicate_min-1000,y=max(rarefaction_obj$Species+10),label=replicate_min,angle=90))+
    theme(panel.background = element_blank(),panel.border = element_rect(fill=NA,color="black"))+
  xlab("Number of Sequences")+
  ylab("ASV Count")
  rarefaction_information<-list(rarefaction_obj,rarefaction_plot)
  return(rarefaction_information)

    
}

```

## Rarefy Table Function

The below function takes a rarfied table as input and removes samples that do not have the minimum sampling depth and any ASV that has a sum across samples of zeros following rarefaction. The function takes as input the following items: 

1. **rarefied_table**: a rarefied ASV table with unrarefied samples and zero sum ASVs still present
2. **sample**: sequencing depth by which all samples were rarefied.

```{r}
# Removing unrarefied samples (those less than the minimum sequencing depth). Removing ASVs present in the ASV table that were only present in samples removed following rarefaction. These ASVs will have column sums of 0. 
rarefy_formatting<-function(nonrarefied_asv_tab,sample){
  rarefied_table<-suppressWarnings(as.data.frame(rrarefy(nonrarefied_asv_tab, sample=sample)))
  ds.rarefied<-as.data.frame(subset(rarefied_table, rowSums(rarefied_table)>=sample))
  ds.rare.asv<-ds.rarefied[,colSums(ds.rarefied)>0]
  if(all(rowSums(ds.rare.asv)==sample)){
    print(paste0("All samples rarefied to ",sample))
  }
  return(ds.rare.asv)
}
```

## PCoA and Ordination Function

The below function wraps the vegdist function from vegan and the pcoa function from ape and generates an ordination using ggplot. The results of the PCoA and the ggplot are stored in a two object list. The first is the PCoA output and the second is the ggplot that can be further customized using ggplot arguments. The function requires the following input:

1. **table**: a rarefied ASV table
2. **method**: a method for computing distances (see: ??vegan::vegdist for options)
3. **point_ellipse_color**: The variable by which points and ellipses should be colored
4. **shape**: Shaping variable
5. **polygon**: Logical value indiciating whether filled ellipses should be plotted to reprsented. This is useful when working with data with multiple factors. **Must be specified with ellipse_fill has a value.** 
6. **ellipse_fill**: Variable by which filled ellipses should be colored. 
7. **point_size**: Size of points for ordination. Default is 3.5

```{r}
pcoa_imager<-function(table,method,point_ellipse_color,shape=NULL,polygon=NULL,ellipse_fill=NULL,point_size=3.5){
  column_grabber<-`[[` # Generating a subsetting function
  point_ellipse_color<-column_grabber(table,point_ellipse_color) # will subset the column that contains the provided variable by which points and ellipses should be colored
  ellipse_fill<-column_grabber(table,ellipse_fill) # will subset the column that contains the variabe by which ellipses will be filled
  shape<-column_grabber(table,shape) # will subset the column that contains the point shaping variabe 
  relabund.asv<-decostand(Filter(x=table,f=is.numeric), method="total") # using the decostand function from vegan to convert count data to relative abundances
  dist.asv<-vegdist(relabund.asv, method=method) # calculating distance matrix using the vegdist function from vegan. Method is a user specified variable to indicate which method for computing distance to be used. 
  pcoa.asv<-pcoa(dist.asv) # using the pcoa function from ape to perform a PCoA with the distance matrix stored in dist.asv. 
  pcoavec.asv<-as.data.frame(pcoa.asv$vectors) # creating a dataframe that only includes the vectors from the PCoA.
  pcoasitescores.asv<-data.frame(PC1=pcoavec.asv$Axis.1, PC2=pcoavec.asv$Axis.2) # creating a datatable that is only the first and second PCs. 
  
  # creating a if/else statement based on presence of ellipse_fill. 
      if (is.null(ellipse_fill)){
      pcoagraph.asv<-data.frame(pcoasitescores.asv,PC1=pcoasitescores.asv$PC1, PC2=pcoasitescores.asv$PC2, color=point_ellipse_color, shape_var=shape)
    } else {
  pcoagraph.asv<-data.frame(pcoasitescores.asv,PC1=pcoasitescores.asv$PC1, PC2=pcoasitescores.asv$PC2, color=point_ellipse_color, ellipse_fill=ellipse_fill, shape_var=shape)}
  
  # Generates standard deviation ellipses based on color variable
  pcoaellipse<-ordiellipse(pcoasitescores.asv,pcoagraph.asv$color, display="sites", kind="sd", draw="none")
  ell <- data.frame()
for(g in levels(as.factor(pcoagraph.asv$color))){
ell <- rbind(ell, cbind(as.data.frame(with(pcoagraph.asv[pcoagraph.asv$color==g,],                                                vegan:::veganCovEllipse(pcoaellipse[[g]]$cov,pcoaellipse[[g]]$center,pcoaellipse[[g]]$scale))) ,color=g))}
  
  # Generates secondary ellipse for second color
pcoaellipse.2<-ordiellipse(pcoasitescores.asv,pcoagraph.asv$ellipse_fill, display="sites", kind="sd", draw="none")
  ell_2 <- data.frame()
for(g in levels(as.factor(pcoagraph.asv$ellipse_fill))){
ell_2 <- rbind(ell_2, cbind(as.data.frame(with(pcoagraph.asv[pcoagraph.asv$elllipse_fill==g,],                                                vegan:::veganCovEllipse(pcoaellipse.2[[g]]$cov,pcoaellipse.2[[g]]$center,pcoaellipse.2[[g]]$scale))) ,ellipse_fill=g))}  
  
# Plotting first two PCs of PCoA in ggplot  
  
  # If an ellipse_fill variable is present polygon must be set to true
  if (polygon==TRUE){
  # If statement that indicates what to do in the absence of a shape variable
    if (is.null(shape)){
      pcoa_plot<-ggplot(pcoagraph.asv, aes(PC1,PC2, colour=color))+
      geom_point(size=point_size)+
      geom_path(data=ell, aes(x=PC1, y=PC2, colour=color),linewidth=0.5, linetype=1)+
  geom_polygon(data=ell_2, aes(x=PC1, y=PC2, fill=ellipse_fill),linewidth=0.5, linetype=2,inherit.aes = FALSE,alpha=0.1)+
  theme(axis.title.x=element_text(size=14, face="bold"))+
  theme(axis.title.y=element_text(size=14, face="bold"))+
  theme(axis.text.x=element_text(size=12, face="bold"))+
  theme(axis.text.y=element_text(size=12, face="bold"))+ 
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
    panel.border=element_rect(colour="black", linewidth =1, fill=NA),
    panel.background = element_rect(fill="white"))+
    xlab(paste0("PC1"," (",round(column_grabber(pcoa.asv,"values")[1,2]*100,2),"%)"))+
    ylab(paste0("PC2"," (",round(column_grabber(pcoa.asv,"values")[2,2]*100,2),"%)"))
   
       } else {
  pcoa_plot<-ggplot(pcoagraph.asv, aes(PC1,PC2, colour=color))+       
  geom_point(aes(shape=shape_var),size=point_size)+
  geom_path(data=ell, aes(x=PC1, y=PC2, colour=color),linewidth=0.5, linetype=1)+
  geom_polygon(data=ell_2, aes(x=PC1, y=PC2, fill=ellipse_fill),linewidth=0.5, linetype=2,inherit.aes = FALSE,alpha=0.1)+
  theme(axis.title.x=element_text(size=14, face="bold"))+
  theme(axis.title.y=element_text(size=14, face="bold"))+
  theme(axis.text.x=element_text(size=12, face="bold"))+
  theme(axis.text.y=element_text(size=12, face="bold"))+ 
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
    panel.border=element_rect(colour="black", linewidth =1, fill=NA),
    panel.background = element_rect(fill="white"))+
    xlab(paste0("PC1"," (",round(column_grabber(pcoa.asv,"values")[1,2]*100,2),"%)"))+
    ylab(paste0("PC2"," (",round(column_grabber(pcoa.asv,"values")[2,2]*100,2),"%)"))}
         
    } else {
      
      if (is.null(shape)){
      pcoa_plot<-ggplot(pcoagraph.asv, aes(PC1,PC2, colour=color))+
      geom_point(size=point_size)+
      geom_path(data=ell, aes(x=PC1, y=PC2, colour=color),linewidth=0.5, linetype=1)+
  theme(axis.title.x=element_text(size=14, face="bold"))+
  theme(axis.title.y=element_text(size=14, face="bold"))+
  theme(axis.text.x=element_text(size=12, face="bold"))+
  theme(axis.text.y=element_text(size=12, face="bold"))+ 
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
    panel.border=element_rect(colour="black", linewidth =1, fill=NA),
    panel.background = element_rect(fill="white"))+
    xlab(paste0("PC1"," (",round(column_grabber(pcoa.asv,"values")[1,2]*100,2),"%)"))+
    ylab(paste0("PC2"," (",round(column_grabber(pcoa.asv,"values")[2,2]*100,2),"%)"))
      
        } else {
 pcoa_plot<-ggplot(pcoagraph.asv, aes(PC1,PC2, colour=color))+ 
  geom_point(aes(shape=shape_var), size=point_size)+
  geom_path(data=ell, aes(x=PC1, y=PC2, colour=color),linewidth=0.5, linetype=1)+
 theme(axis.title.x=element_text(size=14, face="bold"))+
  theme(axis.title.y=element_text(size=14, face="bold"))+
  theme(axis.text.x=element_text(size=12, face="bold"))+
  theme(axis.text.y=element_text(size=12, face="bold"))+ 
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
    panel.border=element_rect(colour="black", size=1, fill=NA),
    panel.background = element_rect(fill="white"))+
      xlab(paste0("PC1"," (",round(column_grabber(pcoa.asv,"values")[1,2]*100,2),"%)"))+
    ylab(paste0("PC2"," (",round(column_grabber(pcoa.asv,"values")[2,2]*100,2),"%)"))}
    }
  pcoa_func_out<-list(pcoa.asv,pcoa_plot)
  return(pcoa_func_out)
}

```

## Taxonomy String Cleaner

A function to clean-up taxonomy string names so that they are more aesthetically pleasing. Function takes as input the following object:

**taxonomy.table**: A taxonomy table object. 

```{r}
taxonomy_string_clean<-function(taxonomy.table){
    taxonomy.table[["Kingdom"]]<-sub("k__","",taxonomy.table[["Kingdom"]])
    taxonomy.table[["Phylum"]]<-sub("p__","",taxonomy.table[["Phylum"]])
    taxonomy.table[["Class"]]<-sub("c__","",taxonomy.table[["Class"]])
    taxonomy.table[["Order"]]<-sub("o__","",taxonomy.table[["Order"]])
    taxonomy.table[["Family"]]<-sub("f__","",taxonomy.table[["Family"]])
    taxonomy.table[["Genus"]]<-sub("g__","",taxonomy.table[["Genus"]])
    taxonomy.table[["Species"]]<-sub("s__","",taxonomy.table[["Species"]])
    taxonomy.table[is.na(taxonomy.table)]<-"Unassigned"
    return(taxonomy.table)
  }
```


## Relative Abundance Plot Function

A function to generates relative abundance bar charts that requires the following input:

1. **metadata**: The object containing metadata info to be used for grouping data. **ROW NAMES MUST MATCH THOSE IN ASV TABLE**
2. **taxonomy.table**: Taxonomy table generated by the dada2 pipeline
3. **asv.table**: Rarefied ASV table
4. **grouping.var**: Grouping variable present in the metadata object 
5. **taxon_level**: Taxonomy level to be used for plotting
6. **cut_off**: Relative abundance cut-off to be shown in the chart

```{r}

relative_abundance_plots<-function(metadata,taxonomy.table,asv.table,grouping.var,taxon_level,cut_off,kingdom){
if(deparse(substitute(taxon_level))=="Kingdom"){
  stop("YOU SPECIFIED TAXON LEVEL AS KINGDOM, THIS WILL BREAK THE CODE AND GIVE YOU A HEADACHE SO STOP")
}
  if(kingdom=="fungi"){
    taxonomy.table[["Kingdom"]]<-sub("k__","",taxonomy.table[["Kingdom"]])
    taxonomy.table[["Phylum"]]<-sub("p__","",taxonomy.table[["Phylum"]])
    taxonomy.table[["Class"]]<-sub("c__","",taxonomy.table[["Class"]])
    taxonomy.table[["Order"]]<-sub("o__","",taxonomy.table[["Order"]])
    taxonomy.table[["Family"]]<-sub("f__","",taxonomy.table[["Family"]])
    taxonomy.table[["Genus"]]<-sub("g__","",taxonomy.table[["Genus"]])
    taxonomy.table[["Species"]]<-sub("s__","",taxonomy.table[["Species"]])
     taxonomy.table[is.na(taxonomy.table)]<-"Unassigned"
  }
  metadata.asv.tab<-merge(metadata,asv.table,by.x=0,by.y=0) # Merges asv table and metadata file by row names
  metadata.asv.tab_group_sums<-metadata.asv.tab[,c(which(colnames(metadata.asv.tab)==paste0("",deparse(substitute(grouping.var)),"")),which(sapply(metadata.asv.tab,is.numeric)))]%>% # Takes the sum of each ASV by the grouping variable
    as.data.frame()%>%
    group_by({{grouping.var}})%>%
    summarize_all(sum)%>%
    as.data.frame()
  row.names(metadata.asv.tab_group_sums)<-metadata.asv.tab_group_sums[[deparse(substitute(grouping.var))]] # Converts grouping variable to rownames
  metadata.asv.tab_group_sums[[deparse(substitute(grouping.var))]]<-NULL # Removes duplicate column 
  t.metadata.asv.tab_group_sums<-as.data.frame(t(metadata.asv.tab_group_sums)) # Transposes group sum data frame
 t.metadata.asv.tab_group_sums[]<-as.data.frame(sapply(t.metadata.asv.tab_group_sums,as.numeric)) # Converts data to numeric
  metadata.asv.tab.taxonomy_group_sums<-merge(taxonomy.table,t.metadata.asv.tab_group_sums,by=0)
  # Merges taxonomy information into group summed ASV table
 metadata.asv.tab.taxonomy_taxon_sums<-metadata.asv.tab.taxonomy_group_sums[,c(which(colnames(metadata.asv.tab.taxonomy_group_sums)==paste0("",deparse(substitute(taxon_level)),"")),which(sapply(metadata.asv.tab.taxonomy_group_sums,is.numeric)))]%>% # Takes the sum of ASVs by their taxonomic assignment
 as.data.frame()%>%
group_by({{taxon_level}})%>%
summarize_all(sum)%>%
as.data.frame()

 metadata.asv.tab.taxonomy_taxon_sums[[deparse(substitute(taxon_level))]][which(is.na(metadata.asv.tab.taxonomy_taxon_sums[[deparse(substitute(taxon_level))]]))]<-"Unassigned" # Renames NA assignments as unassigned 
  rownames(metadata.asv.tab.taxonomy_taxon_sums)<-metadata.asv.tab.taxonomy_taxon_sums[[deparse(substitute(taxon_level))]] # Updates row names to taxonomy names
  metadata.asv.tab.taxonomy_taxon_sums[[deparse(substitute(taxon_level))]]<-NULL # Removes duplicate column
  t.metadata.asv.tab.taxonomy_taxon_sums<-as.data.frame(t(metadata.asv.tab.taxonomy_taxon_sums)) # Transposes data frame 
  t.metadata.asv.tab.taxonomy_taxon_cutoff<-t.metadata.asv.tab.taxonomy_taxon_sums[,which((colSums(t.metadata.asv.tab.taxonomy_taxon_sums[,grep("Unassigned",colnames(t.metadata.asv.tab.taxonomy_taxon_sums),invert=TRUE)])/sum(t.metadata.asv.tab.taxonomy_taxon_sums))>substitute(cut_off))] # Creates an object that has only those ASVs that have a relative abundance greater than the specified cut-off
    t.metadata.asv.tab.taxonomy_taxon_other<-t.metadata.asv.tab.taxonomy_taxon_sums[,which((colSums(t.metadata.asv.tab.taxonomy_taxon_sums[,grep("Unassigned",colnames(t.metadata.asv.tab.taxonomy_taxon_sums),invert=TRUE)])/sum(t.metadata.asv.tab.taxonomy_taxon_sums))<substitute(cut_off))] # Creates an object that contains only the taxa who fell below the specified relative abundance cut-off
  taxon_unassigned<-data.frame(Unassigned=t.metadata.asv.tab.taxonomy_taxon_sums[,grep("Unassigned",colnames(t.metadata.asv.tab.taxonomy_taxon_sums),invert=FALSE)]) # Creates an object to store the unassigned taxa
  other<-data.frame(Other=rowSums( t.metadata.asv.tab.taxonomy_taxon_other)) # Takes the sum of the taxa that fell below the cut-off
  t.metadata.asv.tab.taxonomy_taxon_cutoff_relabund<-decostand(cbind(t.metadata.asv.tab.taxonomy_taxon_cutoff,other,taxon_unassigned),method="total") # Calculates relative abundance 
  t.metadata.asv.tab.taxonomy_taxon_cutoff_relabund_long<-t.metadata.asv.tab.taxonomy_taxon_cutoff_relabund%>%
   tibble::rownames_to_column(var="Treatment")%>%
   pivot_longer(cols=!Treatment,names_to = "taxon_level_plot", values_to = "Relative.Abundance")%>%
    as.data.frame() # Converts data frame to long
  
t.metadata.asv.tab.taxonomy_taxon_cutoff_relabund_long[["taxon_level_plot"]]<-factor(unique(t.metadata.asv.tab.taxonomy_taxon_cutoff_relabund_long[["taxon_level_plot"]]),levels=unique(t.metadata.asv.tab.taxonomy_taxon_cutoff_relabund_long[["taxon_level_plot"]])) # Converts 

colour_count<-length(unique(t.metadata.asv.tab.taxonomy_taxon_cutoff_relabund_long[["taxon_level_plot"]]))
getPalette <- colorRampPalette(RColorBrewer::brewer.pal(9,"Set1"))
  
return(ggplot(t.metadata.asv.tab.taxonomy_taxon_cutoff_relabund_long,aes(x=Treatment,y=Relative.Abundance))+
geom_col(aes(x=Treatment,y=Relative.Abundance,fill=taxon_level_plot),linewidth=0.75,inherit.aes = FALSE,color="black")+
theme(panel.grid=element_blank(),panel.border = element_rect(color="black",fill=NA),panel.background = element_rect(fill="white"),legend.position = "bottom")+
  scale_fill_manual(values = c(getPalette(colour_count-1),"lightgrey"))+
guides(fill=guide_legend(title = deparse(substitute(taxon_level)))))
}

```

## Funguild Formatting Function

A function to generate funguild compatible OTU tables

1. **t_asv_taxonomy_table**: The imported ASV table that has ASV IDs as the rows and sample names as the columns. Should also have taxonomy columns from UNITE (Kingdom through Species). 

```{r}

funguild_formatting<-function(t_asv_taxonomy_table){
  colnames(t_asv_taxonomy_table)[1]<-"OTU ID" # Changes the name of the first column to OTU ID
  t_asv_taxonomy_table$taxonomy<-paste0(t_asv_taxonomy_table$Kingdom,";",t_asv_taxonomy_table$Phylum,";",t_asv_taxonomy_table$Class,";",t_asv_taxonomy_table$Order,";",t_asv_taxonomy_table$Family,";",t_asv_taxonomy_table$Genus,";",t_asv_taxonomy_table$Species) # Creates a column called taxonomy that contains the concatenated taxonomy strings
  t_asv_taxonomy_table_formatted<-subset(t_asv_taxonomy_table,select=-c(Kingdom,Phylum,Class,Order,Family,Genus,Species)) # drops the individual taxonomy columns for each taxonomic level. 
  print(colnames(t_asv_taxonomy_table_formatted))
  return(t_asv_taxonomy_table_formatted)
}

```

# Data Importing

```{r}
t.seqtab.tax_its_soil_roots<-read.table("/Users/5io/Library/CloudStorage/OneDrive-OakRidgeNationalLaboratory/septoria_in_a_bag/amplicon_sequencing/analyses_rdata/04_09_2024_its2_sequence_processing/t.seqtab.tax_its_soil_roots.txt",header=TRUE,sep='\t')

t.seqtab.tax_v4_soil_roots<-read.table("/Users/5io/Library/CloudStorage/OneDrive-OakRidgeNationalLaboratory/septoria_in_a_bag/amplicon_sequencing/analyses_rdata/04_10_2024_v4_sequence_processing/t.seqtab.tax_v4_soil_roots.txt",header=TRUE,sep='\t')

sphaerulina_asv_id<-c("asv_its.384","asv_its.484","asv_its.1326","asv_its.1327","asv_its.1328","asv_its.1329","asv_its.1340","asv_its.1449","asv_its.2754","asv_its.4218")
```


# Data Formatting ITS

```{r}
# Checking the row names. Should be ASV IDs
row.names(t.seqtab.tax_its_soil_roots)

# Checking the column names. Should be sample names
colnames(t.seqtab.tax_its_soil_roots)

# Removing Negatives and EoP from ASV table
t.seqtab.tax_its<-t.seqtab.tax_its_soil_roots[,grep("NTC|Extraction",colnames(t.seqtab.tax_its_soil_roots),invert=TRUE)]

# Subsetting data frame so that only ASV count info is present in ASV table
t.seqtab_its<-t.seqtab.tax_its[,2:(length(t.seqtab.tax_its)-7)]
sum(t.seqtab_its)

# Subsetting data frame so that only taxonomy is present 
tax.tab_its<-t.seqtab.tax_its[,(length(t.seqtab.tax_its)-6):(length(t.seqtab.tax_its))]

# Transposing ASV table back to original format so that it can be used with vegan. 
seqtab_its<-as.data.frame(t(t.seqtab_its))

# Fixing the sample name
row.names(seqtab_its)[59]<-sub("NoSept_NA_P_R2_MoP_D25","NoSept_NA_NP_R2_MoP_D25",row.names(seqtab_its)[59])

```

# Data Formatting v4

```{r}
# Checking the row names. Should be ASV IDs
row.names(t.seqtab.tax_v4_soil_roots)

# Checking the column names. Should be sample names
colnames(t.seqtab.tax_v4_soil_roots)

# Removing Negatives and EoP from ASV table
t.seqtab.tax_v4<-t.seqtab.tax_v4_soil_roots[,grep("NTC|Extraction|Amp|AMP",colnames(t.seqtab.tax_v4_soil_roots),invert=TRUE)]

# Subsetting data frame so that only ASV count info is present in ASV table
t.seqtab_v4<-t.seqtab.tax_v4[,2:(length(t.seqtab.tax_v4)-7)]
sum(t.seqtab_v4)

# Subsetting data frame so that only taxonomy is present 
tax.tab_v4<-t.seqtab.tax_v4[,(length(t.seqtab.tax_v4)-6):(length(t.seqtab.tax_v4))]

# Transposing ASV table back to original format so that it can be used with vegan. 
seqtab_v4<-as.data.frame(t(t.seqtab_v4))

# Fixing the sample name
row.names(seqtab_v4)[59]<-sub("NoSept_NA_P_R2_MoP_D25","NoSept_NA_NP_R2_MoP_D25",row.names(seqtab_v4)[59])

```

# Metadata Object

Generating a metadata object that includes the following information:

1. **Inoculation Type**
  A. SARM07: *Sphaerulina musiva* low virulence
  B. Mn5: *Sphaerulina musiva* high virulence
  C. NoSept: Non-inoculate soils

2. **Autoclave Type**
  A. A: Field soils collected from Oregon that were autoclaved prior to the start of the study and amended into the soil. 
  B. NA: Field soils collected from Oregon that were amended into study soil that were not autoclaved.
  
3. **Plant Presence**
  A. P: *Populus* was present in the pot.
  B. NP: *Populus* was not present in the pot and it was soil only. 
  
4. **Location of Soil Collection**
  A. EoP: Edge of pot 
  B. MoP: Middle of pot
  C. NR: Near root
  
5. **Date of Collection**
  A. D10/D25: Samples with this information present in the beginning of the name were the initial samples collected prior to fungal addition. 
  B. o9: Sequencing was completed for these samples October 9th (likely Day 10 samples)
  C. o15: Sequencing was completed for these samples October 15th (likely Day 25 samples)
  
```{r}
# Creating an empty matrix to ad lib a metadata file
s.bag.meta<-matrix(nrow=334,ncol=1)
s.bag.meta<-as.data.frame(s.bag.meta)
row.names(s.bag.meta)<-row.names(seqtab_its)
s.bag.meta$sample_names<-as.data.frame(row.names(seqtab_its))

# Splitting out information about septoria inoculation from sample names
septoria<-c(sapply(strsplit(row.names(seqtab_its[1:192,]),"_"),'[',1),sapply(strsplit(row.names(seqtab_its[193:286,]),"_"),'[',2),sapply(strsplit(row.names(seqtab_its[287:nrow(seqtab_its),]),"_"),'[',1))

# Splitting out information about soil autoclaving from sample names
autoclave<-c(sapply(strsplit(row.names(seqtab_its[1:192,]),"_"),'[',2),sapply(strsplit(row.names(seqtab_its[193:286,]),"_"),'[',3),sapply(strsplit(row.names(seqtab_its[287:nrow(seqtab_its),]),"_"),'[',2))

# Splitting out information about presence of a plant from sample names
plant<-c(sapply(strsplit(row.names(seqtab_its[1:192,]),"_"),'[',3),sapply(strsplit(row.names(seqtab_its[193:286,]),"_"),'[',4),sapply(strsplit(row.names(seqtab_its[287:nrow(seqtab_its),]),"_"),'[',3))

# Splitting out information about reps from sample names
rep<-c(sapply(strsplit(row.names(seqtab_its[1:192,]),"_"),'[',4),sapply(strsplit(row.names(seqtab_its[193:286,]),"_"),'[',5),sapply(strsplit(row.names(seqtab_its[287:nrow(seqtab_its),]),"_"),'[',4))

# Splitting out information regarding sample location collection information
location<-c(sapply(strsplit(row.names(seqtab_its[1:192,]),"_"),'[',5),rep("initial",times=94),sapply(strsplit(row.names(seqtab_its[287:nrow(seqtab_its),]),"_"),'[',5))

# Splitting out information for sampling date
date<-c(sapply(strsplit(row.names(seqtab_its[1:192,]),"_"),'[',6),sapply(strsplit(row.names(seqtab_its[193:286,]),"_"),'[',6),sapply(strsplit(row.names(seqtab_its[287:nrow(seqtab_its),]),"_"),'[',6))

niche<-c(sapply(strsplit(row.names(seqtab_its[1:192,]),"_"),'[',7),sapply(strsplit(row.names(seqtab_its[193:286,]),"_"),'[',7),sapply(strsplit(row.names(seqtab_its[287:nrow(seqtab_its),]),"_"),'[',7))

# Adding the metadata to the metadata sheet
s.bag.meta$septoria<-septoria
s.bag.meta$autoclave<-autoclave
s.bag.meta$plant<-plant
s.bag.meta$location<-location
s.bag.meta$rep<-rep
s.bag.meta$date<-date
s.bag.meta$niche<-niche
s.bag.meta<-s.bag.meta[,-c(1,2)]

# Fixing one of the septoria labels that were spelt a little different from the others.
s.bag.meta$septoria<-sub("noSept","NoSept",s.bag.meta$septoria)
sample.metadata<-s.bag.meta

#sample_metadata$sept_general<-ifelse(grepl("Mn5|SARM07",sample_metadata$septoria)==TRUE,"septoria","control")

# Creating a new category called plant_date that indicates plant treatment and date of collection.
sample.metadata$plant_date<-paste0(sample.metadata$plant,".",sample.metadata$date)

# Creating a new category called plant_date_septoria that indicates plant treatment, date of collection, sphaerulina inoculant.
sample.metadata$date_septoria<-paste0(sample.metadata$date,"_",sample.metadata$septoria)

# Crating a new category called sept_autoclave that indicates sphaerulina inoculant and autoclave treatment. 
sample.metadata$sept_autoclave<-paste0(sample.metadata$septoria,"_",sample.metadata$autoclave)

# Creating a new category called sept_plant that indicates sphaerulina inoculant and plant treatment
sample.metadata$sept_plant<-paste0(sample.metadata$plant,"_",sample.metadata$septoria)

# Creating a new category called sept_plant that indicates sphaerulina inoculant and plant treatment
sample.metadata$plant_sept_autoclave_date<-paste0(sample.metadata$plant,"_",sample.metadata$septoria,"_",sample.metadata$autoclave,"_",sample.metadata$date)

# Checking sample counts. All should have 4 except for D25 A/NP plants which will have 5 to 6 to try and account for low sequencing depth in MoP samples
as.data.frame(merge(seqtab_its,sample.metadata,by.x=0,by.y=0)) %>%
  group_by(plant_sept_autoclave_date,niche) %>%
  dplyr::count() %>%
  print(n=61)

# Removing duplicate sample
seqtab_its<-seqtab_its[grep("B",row.names(seqtab_its),invert=TRUE),]
```

# S. musiva Relative Abundance

```{r}
# Subsetting ASV table to only include S. musiva ASVs
s.musiva_its<-seqtab_its[,colnames(seqtab_its)%in%sphaerulina_asv_id]

# Merging metadata
s.musiva_its_metadata<-merge(sample.metadata,s.musiva_its,by=0)

s.musiva_its_metadata_long<-gather(s.musiva_its_metadata,key="s.musiva.id",value="count",14:23)


# Plotting in ggplot
ggplot(data=s.musiva_its_metadata_long,aes(x=septoria,y=count,fill=s.musiva.id))+
  geom_boxplot()+
  facet_wrap(~date+autoclave+niche,nrow = 3,scales="free_y")


```


# Data Subsetting

```{r}
# First generating a table that is just root data
seqtab_its_roots<-seqtab_its[grep("roots",row.names(seqtab_its)),]
```


# Sample Rarefaction

```{r}
# Determining number of sequences per sample prior to rarefaction
seqtab_its_sample_counts_roots<-as.data.frame(sort(rowSums(seqtab_its_roots)))

# Adding in metadata to look for patterns in sampling depth
seqtab_its_sample_counts_roots_metadata<-merge(sample.metadata,seqtab_its_sample_counts_roots,by=0)

rare.plot_its_roots<-rarefaction_plotter(nonrarefied_asv_tab = seqtab_its_roots,metadata = sample.metadata,vars = autoclave,color = "autoclave")

rare.plot_its_roots[[2]]

# Rarefying the ITS roots ASV table including both the autoclave and non-autoclaved soils
rare.asv.tab_its_roots<-rarefy_formatting(seqtab_its_roots,sample=sort(rowSums(seqtab_its_roots))[2])

# Rarefying the ITS roots ASV table for non-autoclaved soils only
rare.asv.tab_its_roots_nonautoclave<-rarefy_formatting(seqtab_its_roots[grep("NA",row.names(seqtab_its_roots)),],sample=sort(rowSums(seqtab_its_roots[grep("NA",row.names(seqtab_its_roots)),]))[1])

rare.asv.tab_its_roots_autoclave<-rarefy_formatting(seqtab_its_roots[grep("NA",row.names(seqtab_its_roots),invert=TRUE),],sample=sort(rowSums(seqtab_its_roots[grep("NA",row.names(seqtab_its_roots),invert=TRUE),]))[2])


```


# Data Subsetting

```{r}
# First generating a table that is just root data
seqtab_v4_roots<-seqtab_v4[grep("roots",row.names(seqtab_v4)),]

row.names(seqtab_v4_roots)<-sub("16s_","",row.names(seqtab_v4_roots))
```

# Sample Rarefaction

```{r}
# Determining number of sequences per sample prior to rarefaction
seqtab_v4_sample_counts_roots<-as.data.frame(sort(rowSums(seqtab_v4_roots)))

# Adding in metadata to look for patterns in sampling depth
seqtab_v4_sample_counts_roots_metadata<-merge(sample.metadata,seqtab_v4_sample_counts_roots,by=0)

# Constructing rarefaction curves for fungal root communities
rarefy.plot_v4_roots<-rarecurve(seqtab_v4_roots,tidy = TRUE)

rarefy.plot_v4_roots_metadata<-merge(sample.metadata,rarefy.plot_v4_roots,by.x=0,by.y="Site")

# Plotting rarefaction curves in ggplot
rareplot_v4_roots<-ggplot()+
  geom_line(data=rarefy.plot_v4_roots_metadata,aes(x=Sample,y=Species,colour=autoclave,group=Row.names))+
  geom_vline(data=NULL,xintercept=c(10000,12000,15000,27000))+
  #theme(legend.position="none")+
  geom_text(data=NULL,aes(x=9500,y=3500,label="10000",angle=90))+
  geom_text(data=NULL,aes(x=11500,y=3500,label="12000",angle=90))+
  geom_text(data=NULL,aes(x=14500,y=3500,label="15000",angle=90))+
  geom_text(data=NULL,aes(x=26500,y=3500,label="27000",angle=90))
  

rareplot_v4_roots
  
ggarrange(rareplot_its_roots,rareplot_v4_roots,labels=c("ITS2","V4"),nrow=2,common.legend = TRUE)

# Rarefying the v4 roots ASV table including both the autoclave and non-autoclaved soils
rare.asv.tab_v4_roots<-rarefy_formatting(seqtab_v4_roots,sample=sort(rowSums(seqtab_v4_roots))[2])

# Rarefying the v4 roots ASV table for non-autoclaved soils only
rare.asv.tab_v4_roots_nonautoclave<-rarefy_formatting(seqtab_v4_roots[grep("NA",row.names(seqtab_v4_roots)),],sample=sort(rowSums(seqtab_v4_roots[grep("NA",row.names(seqtab_v4_roots)),]))[1])

rare.asv.tab_v4_roots_autoclave<-rarefy_formatting(seqtab_v4_roots[grep("NA",row.names(seqtab_v4_roots),invert=TRUE),],sample=sort(rowSums(seqtab_v4_roots[grep("NA",row.names(seqtab_v4_roots),invert=TRUE),]))[2])

```

