---
title: "04_09_2024_bacterial_sequence_processing_septoria_in_a_bag"
output: html_document
date: "2024-04-09"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Project Summary** 

The following code describes the sequence processing steps used to analyze **v4** data generated for the **septora-in-a-bag** project. Data comes from DNA extracted from soils and surface sterilized roots. Samples were collected from pots that were amended with either a natural forest soil or an autoclaved forest soil. After soil amendments, *Populus trichocarpa* (black cottonwood) was planted in a subset of pots. The two different strains *Septoria musiva* (AKA *Sphaerulina musiva*) were added to pots. One strain was determined to be high virulence and the other was determined to be low virulence. Soils were collected from three different locations within the pots. 

**Treatments**

**Inoculum Type**

1. **No Septoria** == NoSept
2. **High Virulence Septoria** == Mn5
3. **Low Virulence Septoria** == SARM07

**Sampling Date**

1. **Day 10** == D10
2. **Day 25** == D25

**Soil Amendment Type**

1. **Autoclaved/Reduced Soil Community** == A
2. **Not-Autoclaved/Full Community** == NA

**Plant Presence**

1. **No Plant in Pot** == NP
2. **Plant in Pot** == P

**Soil Collection Location**

1. **Near Root** == NR
2. **Middle of Pot** == MoP
3. **Edge of Pot** == EoP

Middle of pot is the same as near root for pots with plants. 

# **Load Packages**

Loading the packages required to complete the sequence processing. 
```{r}
#Load Packages
library(dada2)
library(Biostrings)
library(ShortRead)
library(reticulate)
library(tidyr)
library(dplyr)
```

# **Path Designations**

Designating the paths for the directory that contains the ITS2 sequencing data from the following sequencing runs:

1. : Soil collected prior to start of experiment (AKA Initial)
2. : Soil collected at day 10 
3. : Soil collected at day 25
4. 

```{r}
# Initial

# Creates an object that has the pathway for the sequence directory
soil_v4_path_initial<-"/Users/5io/Documents/septoria_in_a_bag/soil_amplicon_sequencing/v4_data/v4_amplicon_data_for_processing/230413_M02014_Cregger_AC_SEED_Septoriabag_initial_16s_KJWJG"

# Lists the files present in the sequence directory
list.files(soil_v4_path_initial)

# Creates objects for the forward and reverse reads
forwardreads_v4_soil_initial<-sort(list.files(soil_v4_path_initial, pattern="R1_001",full.names=TRUE))
reversereads_v4_soil_initial<-sort(list.files(soil_v4_path_initial, pattern="R2_001",full.names=TRUE))

# Checks number of forward and reverse files. Should be an equal number of forward and reverse files.
length(forwardreads_v4_soil_initial)
length(reversereads_v4_soil_initial)

# Creates an object  that strips away the pathway information leaving behind only the filename. 
filebasename_v4_soil_initial<-basename(forwardreads_v4_soil_initial)

#  Uses the sub function to strip out the general file info to create sample names that are reflective of sample names in metadata sheet. 
basename_v4_soil_initial<-sub("-16S_.*","_initial_soil",filebasename_v4_soil_initial)

# Uses gsub to substitute the - for the _ which is used in the metadata sheet. 
samplenames_v4_soil_initial<-gsub("\\-","_",basename_v4_soil_initial)

unique(samplenames_v4_soil_initial)


# Day 10 (P2)

# Creates an object that has the pathway for the sequence directory
soil_v4_path_p2<-"/Users/5io/Documents/septoria_in_a_bag/soil_amplicon_sequencing/v4_data/v4_amplicon_data_for_processing/221011_Cregger_AC_SEED_Jo-Soil_Final_P2-16s_KJY9J"

# Lists the files present in the sequence directory
list.files(soil_v4_path_p2)

# Creates objects for the forward and reverse reads
forwardreads_v4_soil_p2<-sort(list.files(soil_v4_path_p2, pattern="R1_001",full.names=TRUE))
reversereads_v4_soil_p2<-sort(list.files(soil_v4_path_p2, pattern="R2_001",full.names=TRUE))

# Checks number of forward and reverse files. Should be an equal number of forward and reverse files.
length(forwardreads_v4_soil_p2)
length(reversereads_v4_soil_p2)

# Creates an object  that strips away the pathway information leaving behind only the filename. 
filebasename_v4_soil_p2<-basename(forwardreads_v4_soil_p2)

#  Uses the sub function to strip out the general file info to create sample names that are reflective of sample names in metadata sheet. 
basename_v4_soil_p2<-sub("-16S_.*","_D10_soil",filebasename_v4_soil_p2)

# Uses gsub to substitute the - for the _ which is used in the metadata sheet. 
samplenames_v4_soil_p2<-gsub("\\-","_",basename_v4_soil_p2)

unique(samplenames_v4_soil_p2)

# Day 25 (P1)

# Creates an object that has the pathway for the sequence directory
soil_v4_path_p1<-"/Users/5io/Documents/septoria_in_a_bag/soil_amplicon_sequencing/v4_data/v4_amplicon_data_for_processing/221005_Cregger_AC_SEED_Jo_Soil_Final_P1-16s_KJYGW"

# Lists the files present in the sequence directory
list.files(soil_v4_path_p1)

# Creates objects for the forward and reverse reads
forwardreads_v4_soil_p1<-sort(list.files(soil_v4_path_p1, pattern="R1_001",full.names=TRUE))
reversereads_v4_soil_p1<-sort(list.files(soil_v4_path_p1, pattern="R2_001",full.names=TRUE))

# Checks number of forward and reverse files. Should be an equal number of forward and reverse files.
length(forwardreads_v4_soil_p1)
length(reversereads_v4_soil_p1)

# Creates an object  that strips away the pathway information leaving behind only the filename. 
filebasename_v4_soil_p1<-basename(forwardreads_v4_soil_p1)

#  Uses the sub function to strip out the general file info to create sample names that are reflective of sample names in metadata sheet. 
basename_v4_soil_p1<-sub("-16S_.*","_D25_soil",filebasename_v4_soil_p1)

# Uses gsub to substitute the - for the _ which is used in the metadata sheet. 
samplenames_v4_soil_p1<-gsub("\\-","_",basename_v4_soil_p1)

# Creates an object  that strips away the pathway information leaving behind only the filename. 
filebasename_v4_soil_p1.2<-basename(reversereads_v4_soil_p1)

# Roots (D10 and D25)
# Creating an object that has the pathway for the sequence directory
roots_v4_path<-"/Users/5io/Documents/septoria_in_a_bag/root_amplicon_sequencing/v4_amplicon_data_for_processing"

# Listing the files present in the sequence directory
list.files(roots_v4_path)

# Creating objects for the forward and reverse reads
forwardreads_v4_roots<-sort(list.files(roots_v4_path, pattern="R1_001",full.names=TRUE))
reversereads_v4_roots<-sort(list.files(roots_v4_path, pattern="R2_001",full.names=TRUE))

# Checking number of forward and reverse files. Should be an equal number of forward and reverse files.
length(forwardreads_v4_roots)
length(reversereads_v4_roots)

# Creating an object  that strips away the pathway information leaving behind only the filename.
filebasename_v4_roots<-basename(forwardreads_v4_roots)

basename_v4_roots<-sub("-16s.*","_roots",filebasename_v4_roots)

# Using gsub to substitute the - for the _ which is used in the metadata sheet. 
samplenames_v4_roots<-gsub("\\-","_",basename_v4_roots)

unique(samplenames_v4_roots)
```


# Initial Quality Assessment

Assessing quality of sequences using a qc plot.
```{r}
# Generates quality plots for initial soils. 
forward_qual_plot_v4_soil_initial_raw<-plotQualityProfile(forwardreads_v4_soil_initial,aggregate = TRUE)
forward_qual_plot_v4_soil_initial_raw
reverse_qual_plot_v4_soil_initial_raw<-plotQualityProfile(reversereads_v4_soil_initial,aggregate = TRUE)
reverse_qual_plot_v4_soil_initial_raw

# Generates quality plots for day 10 soils. 
forward_qual_plot_v4_soil_p2_raw<-plotQualityProfile(forwardreads_v4_soil_p2,aggregate = TRUE)
forward_qual_plot_v4_soil_p2_raw
reverse_qual_plot_v4_soil_p2_raw<-plotQualityProfile(reversereads_v4_soil_p2,aggregate = TRUE)
reverse_qual_plot_v4_soil_p2_raw

# Generates quality plots for day 25 soils.
forward_qual_plot_v4_soil_p1_raw<-plotQualityProfile(forwardreads_v4_soil_p1,aggregate = TRUE)
forward_qual_plot_v4_soil_p1_raw
reverse_qual_plot_v4_soil_p1_raw<-plotQualityProfile(reversereads_v4_soil_p1,aggregate = TRUE)
reverse_qual_plot_v4_soil_p1_raw

# Generating quality plots for day 10 and day 25 roots. 
forward_qual_plot_v4_roots_raw<-plotQualityProfile(forwardreads_v4_roots,aggregate = TRUE)
forward_qual_plot_v4_roots_raw
reverse_qual_plot_v4_roots_raw<-plotQualityProfile(reversereads_v4_roots,aggregate = TRUE)
reverse_qual_plot_v4_roots_raw

```

# Initial Quality Filtering

Removing reads with ambiguous bases prior to using cutadapt to remove primers. 

```{r}
# Initial Soils

# Creates two path objects that will store the paths for the filtered forward and reverse reads. 
forward.filtN_v4_soil_initial<-file.path(soil_v4_path_initial,"filtN",basename(forwardreads_v4_soil_initial))
reverse.filtN_v4_soil_initial<-file.path(soil_v4_path_initial,"filtN",basename(reversereads_v4_soil_initial))

# Filters forward and reverse reads using the filterAndTrim function. Specifying maxN=0 will indicate that sequences with at least 1 ambiguous base will be removed from the dataset. 
filterAndTrim(forwardreads_v4_soil_initial,forward.filtN_v4_soil_initial,reversereads_v4_soil_initial,reverse.filtN_v4_soil_initial,maxN=0,multithread = TRUE)

# Creates quality profiles for the pre-filtered forward and reverse reads.
filtn_qualplot_fwd_v4_soil_initial<-plotQualityProfile(forward.filtN_v4_soil_initial, aggregate=TRUE)
filtn_qualplot_fwd_v4_soil_initial
filtn_qualplot_rev_v4_soil_initial<-plotQualityProfile(reverse.filtN_v4_soil_initial, aggregate=TRUE)
filtn_qualplot_rev_v4_soil_initial

# Soils (Day 10)

# Creates two path objects that will store the paths for the filtered forward and reverse reads. 
forward.filtN_v4_soil_p2<-file.path(soil_v4_path_p2,"filtN",basename(forwardreads_v4_soil_p2))
reverse.filtN_v4_soil_p2<-file.path(soil_v4_path_p2,"filtN",basename(reversereads_v4_soil_p2))

# Filters forward and reverse reads using the filterAndTrim function. Specifying maxN=0 will indicate that sequences with at least 1 ambiguous base will be removed from the dataset. 
filterAndTrim(forwardreads_v4_soil_p2,forward.filtN_v4_soil_p2,reversereads_v4_soil_p2,reverse.filtN_v4_soil_p2,maxN=0,multithread = TRUE)

# Creates quality profiles for the pre-filtered forward and reverse reads.
filtn_qualplot_fwd_v4_soil_p2<-plotQualityProfile(forward.filtN_v4_soil_p2, aggregate=TRUE)
filtn_qualplot_fwd_v4_soil_p2
filtn_qualplot_rev_v4_soil_p2<-plotQualityProfile(reverse.filtN_v4_soil_p2, aggregate=TRUE)
filtn_qualplot_rev_v4_soil_p2

# Soils (Day 25)

# Creates two path objects that will store the paths for the filtered forward and reverse reads. 
forward.filtN_v4_soil_p1<-file.path(soil_v4_path_p1,"filtN",basename(forwardreads_v4_soil_p1))
reverse.filtN_v4_soil_p1<-file.path(soil_v4_path_p1,"filtN",basename(reversereads_v4_soil_p1))

# Filters forward and reverse reads using the filterAndTrim function. Specifying maxN=0 will indicate that sequences with at least 1 ambiguous base will be removed from the dataset. 
filterAndTrim(forwardreads_v4_soil_p1,forward.filtN_v4_soil_p1,reversereads_v4_soil_p1,reverse.filtN_v4_soil_p1,maxN=0,multithread = TRUE)

# Creates quality profiles for the pre-filtered forward and reverse reads.
filtn_qualplot_fwd_v4_soil_p1<-plotQualityProfile(forward.filtN_v4_soil_p1, aggregate=TRUE)
filtn_qualplot_fwd_v4_soil_p1
filtn_qualplot_rev_v4_soil_p1<-plotQualityProfile(reverse.filtN_v4_soil_p1, aggregate=TRUE)
filtn_qualplot_rev_v4_soil_p1

# Roots (Day 10 and Day 25)

# Creating two path objects that will store the paths for the filtered forward and reverse reads. 
forward.filtN_v4_roots<-file.path(roots_v4_path,"filtN",basename(forwardreads_v4_roots))
reverse.filtN_v4_roots<-file.path(roots_v4_path,"filtN",basename(reversereads_v4_roots))

# Filtering forward and reverse reads using the filterAndTrim function. Specifying maxN=0 will indicate that sequences with at least 1 ambiguous base will be removed from the dataset. 
filterAndTrim(forwardreads_v4_roots,forward.filtN_v4_roots,reversereads_v4_roots,reverse.filtN_v4_roots,maxN=0,multithread = TRUE)

# Creates quality profiles for the pre-filtered forward and reverse reads.
filtn_qualplot_fwd_v4_roots<-plotQualityProfile(forward.filtN_v4_roots, aggregate=TRUE)
filtn_qualplot_fwd_v4_roots
filtn_qualplot_rev_v4_roots<-plotQualityProfile(reverse.filtN_v4_roots, aggregate=TRUE)
filtn_qualplot_rev_v4_roots
```

# Trimming Primers

Removing primers using cutadapt. 

## Cutadapt Flags and system connection

Creating a function to generate primer flags for cutadapt. The function takes the following arguments:

**amplicon**: The function comes pre-loaded with the most common primer sets used in the lab. This includes *its2_cregger* and *v4_cregger*. The default is *NULL* on the off-chance that another primer set is used. 

**primer_fwd_list**: If amplicon is NULL then this is a list of user specified forward primer(s).

**primer_rev_list**: If amplicon is NULL then this is a list of user specified reverse primer(s). 

```{r}
# Creating the primer_strings function
primer_strings<-function(amplicon=NULL,primer_fwd_list=NULL,primer_rev_list=NULL){
  if(is.null(amplicon)==FALSE && !amplicon=="its2_cregger" && !amplicon=="v4_cregger"){
    stop("amplicon should be its2_cregger or v4_cregger. If using non-cregger lab primers leave the amplicon argument blank and specify primers in the primer_fwd_list and primer_rev_list")
  }
  cutadapt_primer_flags<-list()
  if(is.null(amplicon)==FALSE && amplicon=="its2_cregger"){  
  its3_forward<-c("CATCGATGAAGAACGCAG","CAACGATGAAGAACGCAG","CACCGATGAAGAACGCAG","CATCGATGAAGAACGTAG","CATCGATGAAGAACGTGG","CATCGATGAAGAACGCTG")
    its4_reverse<-c("TCCTSCGCTTATTGATATGC","TCCTCGCCTTATTGATATGC")
    its3_forward_rc<-lapply(its3_forward,dada2::rc)
    its4_reverse_rc<-lapply(its4_reverse,dada2::rc)
    for (i in 1:length(its3_forward)){
      for(x in 1:length(its4_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",its3_forward[i],"...",its4_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(its4_reverse)){
      for(x in 1:length(its3_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",its4_reverse[i],"...",its3_forward_rc[x], sep=''))
      }
    }    
  }else if(is.null(amplicon)==FALSE && amplicon=="v4_cregger"){
    v4_forward<-c("GTGCCAGCMGCCGCGGTAA","GTGCCAGCMGCWGCGGTAA","GTGCCAGCMGCCGCGGTCA","GTGKCAGCMGCCGCGGTAA")
    v4_reverse<-c("GGACTACHVGGGTWTCTAAT")
    v4_forward_rc<-lapply(v4_forward,dada2::rc)
    v4_reverse_rc<-lapply(v4_reverse,dada2::rc)
    for (i in 1:length(v4_forward)){
      for(x in 1:length(v4_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",v4_forward[i],"...",v4_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(v4_reverse)){
      for(x in 1:length(v4_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",v4_reverse[i],"...",v4_forward_rc[x], sep=''))
      }
    } 
  }else if(is.null(amplicon)==TRUE){
    primer_forward<-primer_fwd_list
    primer_reverse<-primer_rev_list
    primer_forward_rc<-lapply(primer_forward,dada2::rc)
    primer_reverse_rc<-lapply(primer_reverse,dada2::rc)
    for (i in 1:length(primer_forward)){
      for(x in 1:length(primer_reverse_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-a"," ", "^",primer_forward[i],"...",primer_reverse_rc[x], sep=''))
      }
    }
   for (i in 1:length(primer_reverse)){
      for(x in 1:length(primer_forward_rc)){
    cutadapt_primer_flags<-append(cutadapt_primer_flags,paste("-A"," ", "^",primer_reverse[i],"...",primer_forward_rc[x], sep=''))
      }
    } 
  }
    return(cutadapt_primer_flags)
}

# Specifies the conda environment to use for cutadapt
use_condaenv("/opt/anaconda3/envs/cutadaptenv")

# Uses the system2 function to pass commands to the shell to run cutadapt from R. Checking to make sure R can find Cutadapt. 
system2("conda",args=c("run", "-n", "cutadaptenv", "cutadapt", "--version"))

```

## Primer Trimming

```{r}
# Specifying primer strings
v4_primer_strings<-primer_strings(amplicon="v4_cregger")

# Trimming Initial Soils
# Creates a directory to store the forward and reverse reads after they have been trimmed. 
path.cut_initial<-file.path(soil_v4_path_initial,"cutadapt")
if(!dir.exists(path.cut_initial)) dir.create(path.cut_initial)

forwardreads.cut_v4_soil_initial<-file.path(path.cut_initial,basename(forwardreads_v4_soil_initial))
reversereads.cut_v4_soil_initial<-file.path(path.cut_initial,basename(reversereads_v4_soil_initial))

# Uses cutadapt to remove the primers from each read. Removing any read that does not have the forward primer or the reverse primer in the correct orientation.
for (i in seq_along(forwardreads_v4_soil_initial)){
  system2("conda",args=c("run -n cutadaptenv cutadapt",
                         v4_primer_strings,
                          "--discard-untrimmed", "--minimum-length 10",
                          "-o",forwardreads.cut_v4_soil_initial[i], "-p",reversereads.cut_v4_soil_initial[i],
                          forward.filtN_v4_soil_initial[i],reverse.filtN_v4_soil_initial[i]),"--cores=12")
}


# Day 10
# Creates a directory to store the forward and reverse reads after they have been trimmed. 
path.cut_p2<-file.path(soil_v4_path_p2,"cutadapt")
if(!dir.exists(path.cut_p2)) dir.create(path.cut_p2)

forwardreads.cut_v4_soil_p2<-file.path(path.cut_p2,basename(forwardreads_v4_soil_p2))
reversereads.cut_v4_soil_p2<-file.path(path.cut_p2,basename(reversereads_v4_soil_p2))

# Uses cutadapt to remove the primers from each read. Removing any read that does not have the forward primer or the reverse primer in the correct orientation.
for (i in seq_along(forwardreads_v4_soil_p2)){
  system2("conda",args=c("run -n cutadaptenv cutadapt",
                  v4_primer_strings,
                          "--discard-untrimmed", "--minimum-length 10",
                          "-o",forwardreads.cut_v4_soil_p2[i], "-p",reversereads.cut_v4_soil_p2[i],
                          forward.filtN_v4_soil_p2[i],reverse.filtN_v4_soil_p2[i]),"--cores=12")
}

# Day 25
# Creates a directory to store the forward and reverse reads after they have been trimmed. 
path.cut_p1<-file.path(soil_v4_path_p1,"cutadapt")
if(!dir.exists(path.cut_p1)) dir.create(path.cut_p1)

forwardreads.cut_v4_soil_p1<-file.path(path.cut_p1,basename(forwardreads_v4_soil_p1))
reversereads.cut_v4_soil_p1<-file.path(path.cut_p1,basename(reversereads_v4_soil_p1))

# Uses cutadapt to remove the primers from each read. Removing any read that does not have the forward primer or the reverse primer in the correct orientation.
for (i in seq_along(forwardreads_v4_soil_p1)){
  system2("conda",args=c("run -n cutadaptenv cutadapt",
                          v4_primer_strings,
                          "--discard-untrimmed", "--minimum-length 10",
                          "-o",forwardreads.cut_v4_soil_p1[i], "-p",reversereads.cut_v4_soil_p1[i],
                          forward.filtN_v4_soil_p1[i],reverse.filtN_v4_soil_p1[i]),"--cores=12")
}

# Creating a directory to store the forward and reverse reads after they have been trimmed. 
path.cut<-file.path(roots_v4_path,"cutadapt")
if(!dir.exists(path.cut)) dir.create(path.cut)

forwardreads.cut_v4_roots<-file.path(path.cut,basename(forwardreads_v4_roots))
reversereads.cut_v4_roots<-file.path(path.cut,basename(reversereads_v4_roots))

# Using cutadapt to remove the primers from each read. Removing any read that does not have the forward primer or the reverse primer in the correct orientation.

for (i in seq_along(forwardreads_v4_roots)){
  system2("conda",args=c("run -n cutadaptenv cutadapt",
                          v4_primer_strings,
                          "--discard-untrimmed", "--minimum-length 10","--report=minimal",
                          "-o",forwardreads.cut_v4_roots[i], "-p",reversereads.cut_v4_roots[i],
                          forward.filtN_v4_roots[i],reverse.filtN_v4_roots[i]),"--cores=12")
}


# Initial Soils

# Creating new pathways for the quality filtered forward and reverse reads. 
filtforward_v4_soil_initial<-file.path(soil_v4_path_initial, "filtered", (paste0(samplenames_v4_soil_initial,"_F_filt.fastq")))
filtreverse_v4_soil_initial<-file.path(soil_v4_path_initial,"filtered",paste0(samplenames_v4_soil_initial,"_R_filt.fastq"))

filtforward_v4_soil_initial[duplicated(filtforward_v4_soil_initial)]<-sub("filt.fastq.gz","v2_filt.fastq",filtforward_v4_soil_initial[duplicated(filtforward_v4_soil_initial)])
filtreverse_v4_soil_initial[duplicated(filtreverse_v4_soil_initial)]<-sub("filt.fastq.gz","v2_filt.fastq",filtreverse_v4_soil_initial[duplicated(filtreverse_v4_soil_initial)])

# Creating quality profiles for the primer trimmed forward and reverse reads. 
primercut_qualplot_fwd_v4_soil_initial<-plotQualityProfile(forwardreads.cut_v4_soil_initial, aggregate=TRUE)
primercut_qualplot_fwd_v4_soil_initial
primercut_qualplot_rev_v4_soil_initial<-plotQualityProfile(reversereads.cut_v4_soil_initial, aggregate=TRUE)
primercut_qualplot_rev_v4_soil_initial

# Quality filtering reads using a maximum expected error threshold of 2 for the forward and reverse reads. 
filter.out_v4_soil_initial<-filterAndTrim(fwd=forwardreads.cut_v4_soil_initial, filt=filtforward_v4_soil_initial, rev=reversereads.cut_v4_soil_initial, filt.rev=filtreverse_v4_soil_initial, maxEE=c(2,2), compress=FALSE,truncLen = 225, multithread = TRUE)
 
# Converting the filter.out_v4_soil_initial object into a data frame. 
filter.out_v4_soil_initial<-as.data.frame(filter.out_v4_soil_initial)

# Checking to see the effects of the quality filtering on per sample read counts. 
filter.out_v4_soil_initial$diffs <- filter.out_v4_soil_initial$reads.in-filter.out_v4_soil_initial$reads.out

# Using the order command to sort the dataframe by the new column (diffs) that created above. 
filter.out_v4_soil_initial[order(filter.out_v4_soil_initial$reads.out),]

filtforward_v4_soil_initial_no_missing<-filtforward_v4_soil_initial[grep("NTC_Amp1_initial_soil_F_filt.fastq|NTC_Amp2_initial_soil_F_filt.fastq",filtforward_v4_soil_initial,invert = TRUE)]
filtreverse_v4_soil_initial_no_missing<-filtreverse_v4_soil_initial[grep("NTC_Amp1_initial_soil_R_filt.fastq|NTC_Amp2_initial_soil_R_filt.fastq",filtreverse_v4_soil_initial,invert = TRUE)]

# Generating quality plots for the forward and reverse reads after quality filtering
filtered_qualplot_fwd_v4_soil_initial<-plotQualityProfile(filtforward_v4_soil_initial_no_missing, aggregate=TRUE)
filtered_qualplot_fwd_v4_soil_initial
filtered_qualplot_rev_v4_soil_initial<-plotQualityProfile(filtreverse_v4_soil_initial_no_missing, aggregate=TRUE)
filtered_qualplot_rev_v4_soil_initial


# Day 10 Soils

# Creating new pathways for the quality filtered forward and reverse reads. 
filtforward_v4_soil_p2<-file.path(soil_v4_path_p2, "filtered", (paste0(samplenames_v4_soil_p2,"_F_filt.fastq")))
filtreverse_v4_soil_p2<-file.path(soil_v4_path_p2,"filtered",paste0(samplenames_v4_soil_p2,"_R_filt.fastq"))

filtforward_v4_soil_p2[duplicated(filtforward_v4_soil_p2)]<-sub("filt.fastq.gz","v2_filt.fastq",filtforward_v4_soil_p2[duplicated(filtforward_v4_soil_p2)])
filtreverse_v4_soil_p2[duplicated(filtreverse_v4_soil_p2)]<-sub("filt.fastq.gz","v2_filt.fastq",filtreverse_v4_soil_p2[duplicated(filtreverse_v4_soil_p2)])

# Creating quality profiles for the primer trimmed forward and reverse reads. 
primercut_qualplot_fwd_v4_soil_p2<-plotQualityProfile(forwardreads.cut_v4_soil_p2, aggregate=TRUE)
primercut_qualplot_fwd_v4_soil_p2
primercut_qualplot_rev_v4_soil_p2<-plotQualityProfile(reversereads.cut_v4_soil_p2, aggregate=TRUE)
primercut_qualplot_rev_v4_soil_p2

# Quality filtering reads using a maximum expected error threshold of 2 for the forward and reverse reads. 
filter.out_v4_soil_p2<-filterAndTrim(fwd=forwardreads.cut_v4_soil_p2, filt=filtforward_v4_soil_p2, rev=reversereads.cut_v4_soil_p2, filt.rev=filtreverse_v4_soil_p2, maxEE=c(2,2), compress=FALSE,truncLen = 225, multithread = TRUE)
 
# Converting the filter.out_v4_soil_p2 object into a data frame. 
filter.out_v4_soil_p2<-as.data.frame(filter.out_v4_soil_p2)

# Checking to see the effects of the quality filtering on per sample read counts. 
filter.out_v4_soil_p2$diffs <- filter.out_v4_soil_p2$reads.in-filter.out_v4_soil_p2$reads.out

# Using the order command to sort the dataframe by the new column (diffs) that created above. 
filter.out_v4_soil_p2[order(filter.out_v4_soil_p2$reads.out),]

# Generating quality plots for the forward and reverse reads after quality filtering
filtered_qualplot_fwd_v4_soil_p2<-plotQualityProfile(filtforward_v4_soil_p2, aggregate=TRUE)
filtered_qualplot_fwd_v4_soil_p2
filtered_qualplot_rev_v4_soil_p2<-plotQualityProfile(filtreverse_v4_soil_p2, aggregate=TRUE)
filtered_qualplot_rev_v4_soil_p2

# Day 25 Soils

# Creating new pathways for the quality filtered forward and reverse reads. 
filtforward_v4_soil_p1<-file.path(soil_v4_path_p1, "filtered", (paste0(samplenames_v4_soil_p1,"_F_filt.fastq")))
filtreverse_v4_soil_p1<-file.path(soil_v4_path_p1,"filtered",paste0(samplenames_v4_soil_p1,"_R_filt.fastq"))

filtforward_v4_soil_p1[duplicated(filtforward_v4_soil_p1)]<-sub("filt.fastq.gz","v2_filt.fastq",filtforward_v4_soil_p1[duplicated(filtforward_v4_soil_p1)])
filtreverse_v4_soil_p1[duplicated(filtreverse_v4_soil_p1)]<-sub("filt.fastq.gz","v2_filt.fastq",filtreverse_v4_soil_p1[duplicated(filtreverse_v4_soil_p1)])

# Creating quality profiles for the primer trimmed forward and reverse reads. 
primercut_qualplot_fwd_v4_soil_p1<-plotQualityProfile(forwardreads.cut_v4_soil_p1, aggregate=TRUE)
primercut_qualplot_fwd_v4_soil_p1
primercut_qualplot_rev_v4_soil_p1<-plotQualityProfile(reversereads.cut_v4_soil_p1, aggregate=TRUE)
primercut_qualplot_rev_v4_soil_p1

# Quality filtering reads using a maximum expected error threshold of 2 for the forward and reverse reads. 
filter.out_v4_soil_p1<-filterAndTrim(fwd=forwardreads.cut_v4_soil_p1, filt=filtforward_v4_soil_p1, rev=reversereads.cut_v4_soil_p1, filt.rev=filtreverse_v4_soil_p1, maxEE=c(2,2), compress=FALSE,truncLen = 225, multithread = TRUE)
 
# Converting the filter.out_v4_soil_p1 object into a data frame. 
filter.out_v4_soil_p1<-as.data.frame(filter.out_v4_soil_p1)

# Checking to see the effects of the quality filtering on per sample read counts. 
filter.out_v4_soil_p1$diffs <- filter.out_v4_soil_p1$reads.in-filter.out_v4_soil_p1$reads.out

# Using the order command to sort the dataframe by the new column (diffs) that created above. 
filter.out_v4_soil_p1[order(filter.out_v4_soil_p1$reads.out),]

filtforward_v4_soil_p1_no_missing<-filtforward_v4_soil_p1[grep("NTC_2ndAMP_S98_L001_R1_001.fastq.gz_F_filt.fastq",filtforward_v4_soil_p1,invert = TRUE)]
filtreverse_v4_soil_p1_no_missing<-filtreverse_v4_soil_p1[grep("NTC_2ndAMP_S98_L001_R1_001.fastq.gz_R_filt.fastq",filtreverse_v4_soil_p1,invert = TRUE)]

# Generating quality plots for the forward and reverse reads after quality filtering
filtered_qualplot_fwd_v4_soil_p1<-plotQualityProfile(filtforward_v4_soil_p1_no_missing, aggregate=TRUE)
filtered_qualplot_fwd_v4_soil_p1
filtered_qualplot_rev_v4_soil_p1<-plotQualityProfile(filtreverse_v4_soil_p1_no_missing, aggregate=TRUE)
filtered_qualplot_rev_v4_soil_p1



# Roots (Day 10 and Day 25)

# Creating new pathways for the quality filtered forward and reverse reads. 
filtforward_v4_roots<-file.path(roots_v4_path, "filtered", (paste0(samplenames_v4_roots,"_F_filt.fastq")))
filtreverse_v4_roots<-file.path(roots_v4_path,"filtered",paste0(samplenames_v4_roots,"_R_filt.fastq"))

# Creating quality profiles for the primer trimmed forward and reverse reads. 
primercut_qualplot_fwd_v4_roots<-plotQualityProfile(forwardreads.cut_v4_roots, aggregate=TRUE)
primercut_qualplot_fwd_v4_roots
primercut_qualplot_rev_v4_roots<-plotQualityProfile(reversereads.cut_v4_roots, aggregate=TRUE)
primercut_qualplot_rev_v4_roots

# Quality filtering reads using a maximum expected error threshold of 2 for the forward and reverse reads. 
filter.out_v4_roots<-filterAndTrim(fwd=forwardreads.cut_v4_roots, filt=filtforward_v4_roots, rev=reversereads.cut_v4_roots, filt.rev=filtreverse_v4_roots, maxEE=c(2,2), compress=FALSE,truncLen = 225, multithread = TRUE)
 
# Converting the filter.out_v4_soil_p2 object into a data frame. 
filter.out_v4_roots<-as.data.frame(filter.out_v4_roots)

# Checking to see the effects of the quality filtering on per sample read counts. 
filter.out_v4_roots$diffs <- filter.out_v4_roots$reads.in-filter.out_v4_roots$reads.out

# Using the order command to sort the dataframe by the new column (diffs) that created above. 
filter.out_v4_roots[order(filter.out_v4_roots$reads.out),]

# Removing files that were not written as a result of no reads passing filter
filtforward_v4_roots_no_missing<-subset(filtforward_v4_roots,basename(filtforward_v4_roots)%in%list.files(file.path(roots_v4_path, "filtered")))

filtreverse_v4_roots_no_missing<-subset(filtreverse_v4_roots,basename(filtreverse_v4_roots)%in%list.files(file.path(roots_v4_path, "filtered")))

# Generating quality plots for the forward and reverse reads after quality filtering
filtered_qualplot_fwd_v4_roots<-plotQualityProfile(filtforward_v4_roots_no_missing, aggregate=TRUE)
filtered_qualplot_fwd_v4_roots
filtered_qualplot_rev_v4_roots<-plotQualityProfile(filtreverse_v4_roots_no_missing, aggregate=TRUE)
filtered_qualplot_rev_v4_roots
```


## DADA Denoising

Denoising data by first estimating sequencing error rates and using the dada algorithm. 

```{r}
# Initial Soils
# Dereplicating forward and reverse reads
derepforward_v4_soil_initial<-derepFastq(filtforward_v4_soil_initial_no_missing,verbose=TRUE)
derepreverse_v4_soil_initial<-derepFastq(filtreverse_v4_soil_initial_no_missing,verbose=TRUE)

# Assigning the sample names to the dereplicated sequence objects
names(derepforward_v4_soil_initial)<-subset(samplenames_v4_soil_initial,samplenames_v4_soil_initial%in%sub("_F_filt.fastq","",names(derepforward_v4_soil_initial)))
names(derepreverse_v4_soil_initial)<-subset(samplenames_v4_soil_initial,samplenames_v4_soil_initial%in%sub("_F_filt.fastq","",names(derepforward_v4_soil_initial)))

# Learning errors for forward and reverse reads. 
errorforward_v4_soil_initial<-learnErrors(derepforward_v4_soil_initial, multithread = TRUE)
errorreverse_v4_soil_initial<-learnErrors(derepreverse_v4_soil_initial,multithread = TRUE)

# Error plots for forward & reverse reads
errplots_fwd_v4_soil_initial<-plotErrors(errorforward_v4_soil_initial, nominalQ=TRUE)
errplots_fwd_v4_soil_initial
errplots_rev_v4_soil_initial<-plotErrors(errorreverse_v4_soil_initial,nominalQ = TRUE)
errplots_rev_v4_soil_initial

# Denoising with dada2
dadaforwardreads_v4_soil_initial<-dada(derepforward_v4_soil_initial,err=errorforward_v4_soil_initial,multithread = TRUE,pool=TRUE)
dadareversereads.ds_v4_soil_initial<-dada(derepreverse_v4_soil_initial,err=errorreverse_v4_soil_initial,multithread = TRUE,pool=TRUE)

# Day 10 Soils

# Dereplicating forward and reverse reads
derepforward_v4_soil_p2<-derepFastq(filtforward_v4_soil_p2,verbose=TRUE)
derepreverse_v4_soil_p2<-derepFastq(filtreverse_v4_soil_p2,verbose=TRUE)

# Assigning the sample names to the dereplicated sequence objects
names(derepforward_v4_soil_p2)<-subset(samplenames_v4_soil_p2,samplenames_v4_soil_p2%in%sub("_F_filt.fastq","",names(derepforward_v4_soil_p2)))
names(derepreverse_v4_soil_p2)<-subset(samplenames_v4_soil_p2,samplenames_v4_soil_p2%in%sub("_F_filt.fastq","",names(derepforward_v4_soil_p2)))

# Learning errors for forward and reverse reads. 
errorforward_v4_soil_p2<-learnErrors(derepforward_v4_soil_p2, multithread = TRUE)
errorreverse_v4_soil_p2<-learnErrors(derepreverse_v4_soil_p2,multithread = TRUE)

# Error plots for forward & reverse reads
errplots_fwd_v4_soil_p2<-plotErrors(errorforward_v4_soil_p2, nominalQ=TRUE)
errplots_fwd_v4_soil_p2
errplots_rev_v4_soil_p2<-plotErrors(errorreverse_v4_soil_p2,nominalQ = TRUE)
errplots_rev_v4_soil_p2

# Denoising with dada2
dadaforwardreads_v4_soil_p2<-dada(derepforward_v4_soil_p2,err=errorforward_v4_soil_p2,multithread = TRUE,pool=TRUE)
dadareversereads.ds_v4_soil_p2<-dada(derepreverse_v4_soil_p2,err=errorreverse_v4_soil_p2,multithread = TRUE,pool=TRUE)

# Day 25 Soils

# Dereplicating forward and reverse reads
derepforward_v4_soil_p1<-derepFastq(filtforward_v4_soil_p1_no_missing,verbose=TRUE)
derepreverse_v4_soil_p1<-derepFastq(filtreverse_v4_soil_p1_no_missing,verbose=TRUE)

# Assigning the sample names to the dereplicated sequence objects
names(derepforward_v4_soil_p1)<-subset(samplenames_v4_soil_p1,samplenames_v4_soil_p1%in%sub("_F_filt.fastq","",names(derepforward_v4_soil_p1)))
names(derepreverse_v4_soil_p1)<-subset(samplenames_v4_soil_p1,samplenames_v4_soil_p1%in%sub("_F_filt.fastq","",names(derepforward_v4_soil_p1)))

# Learning errors for forward and reverse reads. 
errorforward_v4_soil_p1<-learnErrors(derepforward_v4_soil_p1, multithread = TRUE)
errorreverse_v4_soil_p1<-learnErrors(derepreverse_v4_soil_p1,multithread = TRUE)

# Error plots for forward & reverse reads
errplots_fwd_v4_soil_p1<-plotErrors(errorforward_v4_soil_p1, nominalQ=TRUE)
errplots_fwd_v4_soil_p1
errplots_rev_v4_soil_p1<-plotErrors(errorreverse_v4_soil_p1,nominalQ = TRUE)
errplots_rev_v4_soil_p1


# Denoising with dada2
dadaforwardreads_v4_soil_p1<-dada(derepforward_v4_soil_p1,err=errorforward_v4_soil_p1,multithread = TRUE,pool=TRUE)
dadareversereads.ds_v4_soil_p1<-dada(derepreverse_v4_soil_p1,err=errorreverse_v4_soil_p1,multithread = TRUE,pool=TRUE)



# Roots (Day 10 and Day 25)

# Dereplicating forward and reverse reads
derepforward_v4_roots<-derepFastq(filtforward_v4_roots_no_missing,verbose=TRUE)
derepreverse_v4_roots<-derepFastq(filtreverse_v4_roots_no_missing,verbose=TRUE)

# Assigning the sample names to the dereplicated sequence objects
names(derepforward_v4_roots)<-subset(samplenames_v4_roots,samplenames_v4_roots%in%sub("_F_filt.fastq","",names(derepforward_v4_roots)))
names(derepreverse_v4_roots)<-subset(samplenames_v4_roots,samplenames_v4_roots%in%sub("_F_filt.fastq","",names(derepforward_v4_roots)))

# Learning errors for forward and reverse reads. 
errorforward_v4_roots<-learnErrors(derepforward_v4_roots, multithread = TRUE)
errorreverse_v4_roots<-learnErrors(derepreverse_v4_roots,multithread = TRUE)

# Error plots for forward & reverse reads
errplots_fwd_v4_roots<-plotErrors(errorforward_v4_roots, nominalQ=TRUE)
errplots_fwd_v4_roots
errplots_rev_v4_roots<-plotErrors(errorreverse_v4_roots,nominalQ = TRUE)
errplots_rev_v4_roots

# Denoising forward reads with dada2
dadaforwardreads_v4_roots<-dada(derepforward_v4_roots,err=errorforward_v4_roots,multithread = TRUE,pool=TRUE)

# Denoisng reverse reads with dada2
dadareversereads.ds_v4_roots<-dada(derepreverse_v4_roots,err=errorreverse_v4_roots,multithread = TRUE,pool=TRUE)

# Initial Soils
# Merging forward and reverse reads using mergePairs.
merge_v4_soil_initial<-mergePairs(dadaforwardreads_v4_soil_initial,derepforward_v4_soil_initial,dadareversereads.ds_v4_soil_initial,derepreverse_v4_soil_initial,verbose=TRUE)

# Creating a sequence table
seqtab_v4_soil_initial<-makeSequenceTable(merge_v4_soil_initial)
hist(nchar(colnames(seqtab_v4_soil_initial)))

# Day 10
# Merging forward and reverse reads using mergePairs.
merge_v4_soil_p2<-mergePairs(dadaforwardreads_v4_soil_p2,derepforward_v4_soil_p2,dadareversereads.ds_v4_soil_p2,derepreverse_v4_soil_p2,verbose=TRUE)

# Creating a sequence table
seqtab_v4_soil_p2<-makeSequenceTable(merge_v4_soil_p2)
hist(nchar(colnames(seqtab_v4_soil_p2)))

# Day 25
# Merging forward and reverse reads using mergePairs.
merge_v4_soil_p1<-mergePairs(dadaforwardreads_v4_soil_p1,derepforward_v4_soil_p1,dadareversereads.ds_v4_soil_p1,derepreverse_v4_soil_p1,verbose=TRUE,minOverlap = 12)

# Creating a sequence table
seqtab_v4_soil_p1<-makeSequenceTable(merge_v4_soil_p1)
hist(nchar(colnames(seqtab_v4_soil_p1)))

# Roots Day 10 and Day 25

# Merging forward and reverse reads using mergePairs.
merge_v4_roots<-mergePairs(dadaforwardreads_v4_roots,derepforward_v4_roots,dadareversereads.ds_v4_roots,derepreverse_v4_roots,verbose=TRUE)

# Creating a sequence table
seqtab_v4_roots<-makeSequenceTable(merge_v4_roots)
hist(nchar(colnames(seqtab_v4_roots)))


# Merging ASV tables from all four runs
merged_sequence_table<-mergeSequenceTables(seqtab_v4_soil_p1,seqtab_v4_soil_p2,seqtab_v4_soil_initial,seqtab_v4_roots)

# Removing chimeras using the removeBimeraDenovo function.
seqtab.nochim_v4<-removeBimeraDenovo(merged_sequence_table,method="consensus", multithread=TRUE, verbose=TRUE)

# Number of ASVs in the study
ncol((seqtab.nochim_v4))
nrow(seqtab.nochim_v4)

# Number of sequences
sum(seqtab.nochim_v4)

# Assigning taxonomy using the assignTaxonomy function.
taxa_v4<-assignTaxonomy(seqtab.nochim_v4,"/Users/5io/Documents/septoria_in_a_bag/soil_amplicon_sequencing/v4_data/silva_nr99_v138.1_wSpecies_train_set.fa.gz",multithread = TRUE, minBoot=80)

# Converting the taxonomy assignments to data frame so we can filter the taxonomic assignments.
taxa.original_v4<-as.data.frame(taxa_v4)

# Filtering out ASVs not assigned to a phyla
taxa.na.omit_v4<-taxa.original_v4[-(which(is.na(taxa.original_v4$Phylum))),]

# Filtering out ASVss not assigned to bacteria or archaea
taxa.bacteria_archaea_v4<-taxa.na.omit_v4[grep("Bacteria|Archaea",taxa.na.omit_v4$Kingdom),]

# Filtering out ASVs assingned to a chloroplast
taxa.bacteria_archaea_chloroplast_omit_v4<-taxa.bacteria_archaea_v4[grep("Chloroplast",taxa.bacteria_archaea_v4$Order,invert=TRUE),]

# Filtering out ASVs assingned to a chloroplast
taxa.bacteria_archaea_chloroplast_mitochondria_omit_v4<-taxa.bacteria_archaea_chloroplast_omit_v4[grep("Mitochondria",taxa.bacteria_archaea_chloroplast_omit_v4$Family,invert=TRUE),]

```

## **Taxonomy Merging and ASV Filtering**
The last step is to filter the ASV abundance table by merging the ASV table with the taxonomy assignments. This will remove ASVs from the count table that were not assigned to a fungal phyla or were identified as nematodes or plant sequences. 
```{r}
# Transposing the ASV table so that taxonomy can be added. 
t.seqtab.nochim_v4<-t(seqtab.nochim_v4)

# Merging the two tables together based on row name. 
t.seqtab.nochim.filt_v4<-t.seqtab.nochim_v4[row.names(t.seqtab.nochim_v4)%in%row.names(taxa.bacteria_archaea_chloroplast_mitochondria_omit_v4),]

colnames(t.seqtab.nochim.filt_v4)<-sub("16s_S[0-9].*","16s_roots",colnames(t.seqtab.nochim_v4))

# Number of ASVs & number of sequences post-filtering
nrow(t.seqtab.nochim.filt_v4)
sum(t.seqtab.nochim.filt_v4)

# Merging taxonomy information into ASV table
t.seqtab.tax_v4<-merge(t.seqtab.nochim.filt_v4,taxa.bacteria_archaea_chloroplast_mitochondria_omit_v4, by="row.names")

#  Creating ASV labels and make these new row names.
asvnumber_v4<-as.character(c(1:nrow(t.seqtab.tax_v4)))
asvnumber_v4<-paste0("asv_v4.",labels(asvnumber_v4))

row.names(t.seqtab.tax_v4)<-NULL
row.names(t.seqtab.tax_v4)<-asvnumber_v4


write.table(t.seqtab.tax_v4,"/Users/5io/Library/CloudStorage/OneDrive-OakRidgeNationalLaboratory/septoria_in_a_bag/amplicon_sequencing/analyses/04_10_2024_v4_sequence_processing/t.seqtab.tax_v4_soil_roots.txt",row.names = TRUE,col.names = TRUE,sep='\t')
```


